---
title: "Real Data Analysis"
author: "Ahmed Karam"
date: "2025-01-24"
output:
  pdf_document: default
  html_document: default
---

# Methylprednisolone Data Analysis

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

```{r}
library(dplyr)

library(ggplot2)

library(reshape2)

```

```{r}

library(readxl)

methylprednisolone <- read_excel("methylprednisolone.xlsx", 
    sheet = "Lembar1", 
    range = "A1:AF39")

View(methylprednisolone)

```

# Format Check
```{r}

for_analysis <- methylprednisolone

head(for_analysis)
str(for_analysis)

for_analysis$Group[for_analysis$Group == 1] <- 'methylprednisolone'
for_analysis$Group[for_analysis$Group == 2] <- 'placebo'

for_analysis$Sex[for_analysis$Sex == 1] <- "Male"
for_analysis$Sex[for_analysis$Sex == 2] <- "Female"

for_analysis$`Birth delivered`[for_analysis$`Birth delivered` == 1] <- "C-section"
for_analysis$`Birth delivered`[for_analysis$`Birth delivered` == 2] <- "Spontaneous"

unique(for_analysis$Group)
unique(for_analysis$Sex)
unique(for_analysis$`Birth delivered`)

methylprednisolone_group <- for_analysis$Group == 'methylprednisolone'
placebo_group <- for_analysis$Group == 'placebo'

for_analysis

for_analysis <- for_analysis %>%
  mutate(across(everything(), ~ gsub(",", ".", ., fixed = TRUE)))

```

# 
```{r}

mean_and_sd_info <- for_analysis %>% 
  group_by(Group) %>% 
  summarise(across(`Age (week)`, list(Mean=mean, SD=sd))) %>% 
  mutate(across(2:3, function(x) {round(x, digits = 2)}))

mean_and_sd_info

write.csv(mean_and_sd_info, 'results-1-mean_and_sd_info.csv')

```

# 
```{r}

group_counts <- for_analysis %>% 
  group_by(Group) %>% 
  count(Sex)

group_counts

```

# 
```{r}

group_percentage <- for_analysis %>% 
  group_by(Group) %>% 
  count(Sex) %>% 
  group_by(Group) %>% 
  mutate(across(n, function(x) {x / sum(x) * 100})) %>% 
  mutate(across(n, function(x) {round(x, digits = 2)}))

group_percentage

```

# Basic characteristics of study participants

## Age (Independent sample t-test)

```{r}

colnames(for_analysis)

AGE <- for_analysis %>% 
  select(all_of("Age (week)"))

g1 <- AGE[methylprednisolone_group,]
g2 <- AGE[placebo_group,]

g1_vector <- unlist(g1)
g2_vector <- unlist(g2)

shapiro.test(g1_vector)
shapiro.test(g2_vector)

t.test(g1_vector, g2_vector)

```

## Onset of jaundice (Mann–Whitney Test)

```{r}

colnames(for_analysis)

OFJ <- for_analysis %>% 
  select(all_of("Onset of jaundice (week)"))

g1 <- OFJ[methylprednisolone_group,]
g2 <- OFJ[placebo_group,]

g1_vector <- unlist(g1)
g2_vector <- unlist(g2)

shapiro.test(g1_vector)
shapiro.test(g2_vector)

suppressWarnings(wilcox.test(g1_vector,g2_vector))

```

## Gender (Chi-square tests)

```{r}

colnames(for_analysis)

GENDER <- for_analysis %>% 
  select(all_of(c("Group","Sex")))

tableGENDER <- table(GENDER)

chisq.test(tableGENDER)

```

## Birth delivered

```{r}

colnames(for_analysis)

GENDER <- for_analysis %>% 
  select(all_of(c("Group","Birth delivered")))

tableGENDER <- table(GENDER)

chisq.test(tableGENDER)

```

## Duration of illness (Independent sample t-test)

```{r}

colnames(for_analysis)

DOI <- for_analysis %>% 
  select(all_of("Duration of illness (mingu)"))

g1 <- DOI[methylprednisolone_group,]
g2 <- DOI[placebo_group,]

g1_vector <- unlist(g1)
g2_vector <- unlist(g2)

shapiro.test(g1_vector)
shapiro.test(g2_vector)

t.test(g1_vector, g2_vector)

```

## Birth weight (Mann–Whitney Test)

```{r}

colnames(for_analysis)

BW <- for_analysis %>% 
  select(all_of("Birth weight (gram)"))

g1 <- BW[methylprednisolone_group,]
g2 <- BW[placebo_group,]

g1_vector <- unlist(g1)
g2_vector <- unlist(g2)

shapiro.test(g1_vector)
shapiro.test(g2_vector)

suppressWarnings(wilcox.test(g1_vector,g2_vector))

```

# Basic laboratory measurements before intervention

```{r}

ttest_function <- function(dataFrame, colName, groupOne, groupTwo) {
  
  selectedCols <- dataFrame %>% 
    select(all_of(colName))
  
  g1 <- selectedCols[groupOne,]
  g2 <- selectedCols[groupTwo,]
  
  g1_vector <- as.numeric(unlist(g1))
  g2_vector <- as.numeric(unlist(g2))
  
  output_data_frame <- data.frame(
    Variable = colName,
    
    Shapiro_variable_1_p_value = shapiro.test(g1_vector)$p.value,
    Shapiro_variable_2_p_value = shapiro.test(g2_vector)$p.value,
    
    t_test_p_value = t.test(g1_vector, g2_vector)$p.value
  )
  
  print(output_data_frame)
  
  mmmms <- for_analysis %>% 
  group_by(Group) %>% 
  select(all_of(c(colName))) %>% 
  mutate(across(1,as.numeric)) %>% 
  summarise(across(1,list(Median=median, Min=min, Max=max, Mean=mean, SD=sd)))
  
  print(mmmms)
  }

ttest_function(for_analysis, "Hb", methylprednisolone_group, placebo_group)
ttest_function(for_analysis, "WBC(10^3)", methylprednisolone_group, placebo_group)
ttest_function(for_analysis, "Albumin", methylprednisolone_group, placebo_group)

```

## 

```{r}

wilcox_function <- function(dataFrame, colName, groupOne, groupTwo) {
  
  selectedCols <- dataFrame %>% 
    select(all_of(colName))
  
  g1 <- selectedCols[groupOne,]
  g2 <- selectedCols[groupTwo,]
  
  g1_vector <- as.numeric(unlist(g1))
  g2_vector <- as.numeric(unlist(g2))
  
  output_data_frame <- data.frame(
    Variable = colName,
    
    Shapiro_variable_1_p_value = shapiro.test(g1_vector)$p.value,
    Shapiro_variable_2_p_value = shapiro.test(g2_vector)$p.value,
    
    wilcox_p_value = suppressWarnings(wilcox.test(g1_vector,g2_vector))$p.value
    )
  
  print(output_data_frame)
  
  mmmms <- for_analysis %>% 
    group_by(Group) %>% 
    select(all_of(c(colName))) %>% 
    mutate(across(1,as.numeric)) %>% 
    summarise(across(1,list(Median=median, Min=min, Max=max, Mean=mean, SD=sd)))
  
  print(mmmms)
  }

wilcox_function(for_analysis, "APTT", methylprednisolone_group, placebo_group)
wilcox_function(for_analysis, "PPT", methylprednisolone_group, placebo_group)

```

# Basic laboratory measurements before intervention

```{r}

ttest_paired_function <- function(dataFrame, colNameOne, colNameTwo, theGroup, theGroupName) {
  
  selectedCols <- dataFrame %>% 
    select(all_of(c(colNameOne,colNameTwo)))
  
  g1 <- selectedCols[theGroup,colNameOne]
  g2 <- selectedCols[theGroup,colNameTwo]
  
  g1_vector <- as.numeric(unlist(g1))
  g2_vector <- as.numeric(unlist(g2))
  
  print(selectedCols[theGroup,])
  
  mmmms <- for_analysis %>% 
    group_by(Group) %>% 
    select(all_of(c(colNameOne,colNameTwo))) %>% 
    mutate(across(1:2,as.numeric)) %>% 
    summarise(across(1:2,list(Median=median, Min=min, Max=max, Mean=mean, SD=sd)))
  
  print(mmmms)
  
  boxplot_dataframe <- data.frame(
    variable = c(rep(colNameOne, length(g1_vector)), rep(colNameTwo,length(g2_vector))),
    value = c(g1_vector, g2_vector)
  )
  
  draw_plot <- ggplot(boxplot_dataframe,aes(x=reorder(variable, order(factor(variable), decreasing = T)), y=value)) + 
    geom_boxplot() + geom_jitter() + theme_minimal() + xlab('variable')
  
  print(draw_plot)
  
  output_data_frame <- data.frame(
    Variable_before = colNameOne,
    Variable_after = colNameTwo,
    
    Shapiro_variable_1_p_value = shapiro.test(g1_vector)$p.value,
    Shapiro_variable_2_p_value = shapiro.test(g2_vector)$p.value,
    
    t_test_p_value = t.test(g1_vector,g2_vector,paired=T)$p.value
    )
  
  print(output_data_frame)
  
  write.csv(output_data_frame, sprintf('ttest_paired_function-%s-%s-%s.csv', colNameOne, colNameTwo, theGroupName), row.names=F)
  
}

ttest_paired_function(for_analysis, "Direct bilirubin-pre",
                    "Direct bilirubin-post", placebo_group, "placebo_group")

ttest_paired_function(for_analysis, "Total bilirubin-pre", "Total bilirubin-post", placebo_group, "placebo_group")

ttest_paired_function(for_analysis, "AST-pre", "AST-post", methylprednisolone_group, "methylprednisolone_group")

```

## 

```{r}
wilcox_paired_function <- function(dataFrame, colNameOne, colNameTwo, theGroup, theGroupName) {
  
  selectedCols <- dataFrame %>% 
    select(all_of(c(colNameOne,colNameTwo)))
  
  g1 <- selectedCols[theGroup,colNameOne]
  g2 <- selectedCols[theGroup,colNameTwo]
  
  g1_vector <- as.numeric(unlist(g1))
  g2_vector <- as.numeric(unlist(g2))
  
  print(selectedCols[theGroup,])
  
  mmmms <- for_analysis %>% 
    group_by(Group) %>% 
    select(all_of(c(colNameOne,colNameTwo))) %>% 
    mutate(across(1:2,as.numeric)) %>% 
    summarise(across(1:2,list(Median=median, Min=min, Max=max, Mean=mean, SD=sd)))
  
  print(mmmms)
  
  boxplot_dataframe <- data.frame(
    variable = c(rep(colNameOne, length(g1_vector)), rep(colNameTwo,length(g2_vector))),
    value = c(g1_vector, g2_vector)
  )
  
  draw_plot <- ggplot(boxplot_dataframe,aes(x=reorder(variable, order(factor(variable), decreasing = T)), y=value)) + geom_boxplot() + geom_jitter() + theme_minimal() + xlab('variable')
  
  print(draw_plot)
  
  output_data_frame <- data.frame(
    Variable_before = colNameOne,
    Variable_after = colNameTwo,
    
    Shapiro_variable_1_p_value = shapiro.test(g1_vector)$p.value,
    Shapiro_variable_2_p_value = shapiro.test(g2_vector)$p.value,
    
    Wilcox_test_p_value = suppressWarnings(wilcox.test(g1_vector,g2_vector,paired=T))$p.value
    )
  
  print(output_data_frame)
  
  write.csv(output_data_frame, sprintf('wilcox_paired_function-%s-%s-%s.csv', colNameOne, colNameTwo, theGroupName), row.names=F)
  
}

wilcox_paired_function(for_analysis, "Direct bilirubin-pre", "Direct bilirubin-post", methylprednisolone_group, "methylprednisolone_group")

wilcox_paired_function(for_analysis, "Total bilirubin-pre", "Total bilirubin-post", methylprednisolone_group, "methylprednisolone_group")

wilcox_paired_function(for_analysis, "AST-pre", "AST-post", placebo_group, "placebo_group")

wilcox_paired_function(for_analysis, "GGT-pre", "GGT-post", placebo_group, "placebo_group")

wilcox_paired_function(for_analysis, "GGT-pre", "GGT-post", methylprednisolone_group, "methylprednisolone_group")

```

