---
title: "Sample Size Calculation session"
author: "Hoda Aufy"
date: "2025-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
#Why do we need Sample size calculation?
In Designing an experiment, One of the key questions is:
How many samples do I need to conduct my study?
Too small a sample size: Risks failing to detect the effect of interest.
Too large a sample size: Leads to unnecessary waste of time and resources.

The Goal of Sample Size Calculation:
To determine the optimal number of samples that:
Ensures a reasonable chance of detecting a true effect.
Avoids wasting limited resources on excessive sampling.

#Key Points of Sample Size Calculation:

#Effect Size:
Measures the magnitude of the effect or difference you expect to observe.
The Larger the effect sizes , the easier it is to detect an effect and require fewer samples
#Power (1 - β): 
The probability of detecting a true effect when it exists .
Typically set to 0.8 (80%), meaning there's an 80% chance of detecting the effect.
The higher the power, the more likely it is to detect an effect if present and the more samples needed.
#Significance Level (α):
The threshold for rejecting the null hypothesis .
Commonly set at 0.05, indicating a 5% risk of a false positive. 
 The lower the significance level, the more likely it is to avoid a false positive and the more samples needed.

#Relationship between Sample size calculation and Primary aim of study:
The primary aim of a study plays a crucial role in determining the appropriate sample size, as it directly influences the study design, statistical power, and reliability of results.

#Sample Size calculation using R:

```{r}
install.packages("pwr")
library(pwr)
```

#for comparing 2 means (t-tests):
```{r}
pwr.t.test(d = 0.5, power = 0.8, sig.level = 0.05, type = "two.sample")


```
d= effect size

#For comparing more than 2 means (ANOVA):
```{r}
pwr.anova.test(k = 3, f = 0.25, power = 0.8, sig.level = 0.05)
```
k=Number of groups
f= effect size

# For comparing 2 proportions:
```{r}
power.prop.test(p1=0.45,p2=0.12,sig.level = 0.05,power = 0.8)
```
p1= prevalence of your interest in group 1
p2= prevalence of your interest in group 2

#For correlation studies:
```{r}
pwr.r.test(r = 0.3, power = 0.8, sig.level = 0.05)
```
r= Expected correlation coffecient

#For questionnaire (survey) based studies:
```{r}
install.packages("survey")
library(survey)
z= 1.96
p= 0.5
e=0.05
n <- (z^2 * p * (1 - p)) / e^2
ceiling(n)
```

z= z-score #z=1.96 if your confidence level is 95%
p= estimated proportions
e= marrgin error

#Sample Size for Diagnostic tests:
```{r}
install.packages("SampleSizeDiagnostics")
library(SampleSizeDiagnostics)
sn=0.9
sp=0.85
p=0.2

SampleSizeDiagnostics(sn, sp, p, w = 0.1, CI = 0.95)
```

sn= sensitivity of diagnostic test
sp=specificity of diagnostic test
p= prevalence of disease
w= desired width of confidence interval (default is 0.1) ## unless it is mentioned in the reference.
CI= confidence interval level.



