### Summary of Operators in R

#### 1. **Relational Operators**:
These operators help in comparing values and return a logical (Boolean) result (TRUE or FALSE).

- `==`: Checks equality.
- `!=`: Checks inequality.
- `>`: Checks if the left value is greater than the right.
- `>=`: Checks if the left value is greater than or equal to the right.
- `<`: Checks if the left value is less than the right.
- `<=`: Checks if the left value is less than or equal to the right.

Example:
```{r}
var_1 = c(1, 2, 3, 10, 11)
var_2 = 10
var_1 == var_2   # FALSE FALSE FALSE TRUE FALSE
var_1[var_1 == var_2]  # 10
```

#### 2. **Logical Operators**:
These operators combine multiple logical conditions and return logical values.

- `&`: Logical AND.
- `|`: Logical OR.
- `!`: Logical NOT.

Example:
```{r}
var_3 <- 10
var_3 == var_2 & var_3 >= var_2  # TRUE
var_3 == var_2 & var_3 > var_2   # FALSE
```

#### 3. **Membership Operator**:
Checks if a value is present in a vector or not.

- `%in%`: Checks if an element exists in a vector.

Example:
```{r}
vector_1 <- c(1:5)
value_a <- 4
value_b <- 6
value_a %in% vector_1  # TRUE
value_b %in% vector_1  # FALSE
```

These operators are fundamental for making decisions and performing conditional operations in R.
```

When you save and render this R Markdown notebook, it will produce an HTML document containing the summary and code examples along with any plots or outputs.