---
title: "R recap"
author: "Nada"
date: "2025-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(psych)
library(dplyr)
library(ggplot2)
library(reshape2)
library(ggpubr)
library(grid)
library(vcd)
library(magrittr) 
library(car)
library(plyr)
library(RColorBrewer)
library(gplots)
library(GGally)
```

```{r}
data = read.csv("Cancer_Prediction_Dataset_for _Model_1.csv")
```


Three time-point dataset of pancreatic cancer patients and control group with their clinical data and metabolite expression

```{r}
head(data)
```

What are the summary statistics for each variable (age, BMI, HbA1c_level, blood_glucose_level)?

```{r}
summary(data)
```
```{r}
dim(data)
```



>> What are the counts or percentages for categorical variables (gender, hypertension, heart_disease, smoking_history)?

```{r}
# Gender distribution
table(data$gender)
prop.table(table(data$gender)) * 100
```
```{r}
# Hypertension distribution
table(data$hypertension)
prop.table(table(data$hypertension)) * 100

```


```{r}
# Heart disease distribution
table(data$heart_disease)
prop.table(table(data$heart_disease)) * 100
```


```{r}
# Smoking history
table(data$smoking_history)
prop.table(table(data$smoking_history)) * 100

```



>> the distribution of the Pancreatic Cancer patients according to smoking_history

```{r}
t1 = table(data$smoking_history,data$Pancreatic_Cancer)
prop.table(t1)
```


>> checking for the outliers and are they really outliers?

```{r}
melted <- melt(data[, c("age", "bmi", "HbA1c_level", "blood_glucose_level")], variable.name = "Variable", value.name = "Value")
```

```{r}
ggplot(melted, aes(x = Variable, y = log(Value), fill = Variable)) +
  geom_boxplot( alpha = 0.7) + 
  geom_jitter(width = 0.2, color = "black", alpha = 0.5, size = 1) +  
  labs(
    title = "Boxplot of clinical data ",
    y = "Values",
    fill = "Variables"
  ) + 
  theme_minimal() +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
  ) +
  scale_fill_brewer(palette = "Pastel1") +  
  coord_flip() 
```


>> is there any relation between the clinical data and the Pancreatic_Cancer?

```{r}
# Gender and Pancreatic_Cancer
table(data$gender, data$Pancreatic_Cancer)
chisq.test(table(data$gender, data$Pancreatic_Cancer))
```

```{r}
# Hypertension and Pancreatic_Cancer
table(data$hypertension, data$Pancreatic_Cancer)
chisq.test(table(data$hypertension, data$Pancreatic_Cancer))
```

```{r}
# glucose level and Pancreatic_Cancer
table(data$blood_glucose_level, data$Pancreatic_Cancer)
chisq.test(table(data$blood_glucose_level, data$Pancreatic_Cancer))
```

```{r}
# HbA1c_level and Pancreatic_Cancer
table(data$HbA1c_level, data$Pancreatic_Cancer)
chisq.test(table(data$HbA1c_level, data$Pancreatic_Cancer))

```


# 0 to 0.1: Weak association
# 0.1 to 0.3: Moderate association
# > 0.3: Strong association


>> checking the strength  of the relation between them

```{r}
assocstats(table(data$blood_glucose_level, data$Pancreatic_Cancer))
```
```{r}
assocstats(table(data$HbA1c_level, data$Pancreatic_Cancer))
```



```{r}
rownames(data) = data$Sample
data$Sample = NULL 
```


>> normalize the expression value for the metabolites with median normalization 

```{r}
median_normalization = apply(data[,10:length(data)], 1, function(x) {
  med <- median(x, na.rm = TRUE)  
  x / med                       
})
```

```{r}
CTRL = as.data.frame(median_normalization) %>% select(contains('ctrl')) %>% colMeans() %>% shapiro.test
D0 = as.data.frame(median_normalization) %>% select(contains('D0')) %>% colMeans() %>% shapiro.test
D15 = as.data.frame(median_normalization) %>% select(contains('D15')) %>% colMeans() %>% shapiro.test
D30 = as.data.frame(median_normalization) %>% select(contains('D30')) %>% colMeans() %>% shapiro.test
CTRL
D0
D15
D30
```

>> Non-parametric 


```{r}
qqnorm(median_normalization, main = "QQ Plot of normalized metabolites expression")
```

```{r}
heatmap( median_normalization )
```

```{r}
scaled = scale(median_normalization)
head(scaled)
```

```{r}
qqnorm(scaled, main = "QQ Plot of the scaled normalized metabolites expression")
```


>> prepossessing the expression of the metabolites 

re-merging the clinical data and the expression values to filter samples with NAs  

```{r}
processed_data = merge(data[,1:9],t(scaled), by = "row.names")
```

```{r}
rownames(processed_data) = processed_data$Row.names
processed_data$Row.names = NULL
```


```{r}
rowSums(is.na(processed_data))
```


in Control group there are missing values in 2 samples 22 and 25 
in Do there are no missing values BUT 
in D15 there are 2 samples with missing values 6 and 18 and in D30 there are 3 samples with missing values 11, 21 and 22 
So we need to exclude these samples from our analysis..


>> samples (6,18,11,21,22,25) will be deleted

1- creating two new columns Sample label and Sample Number.

```{r}
processed_data$sample_number = sapply(strsplit(rownames(processed_data), "_"), function(x) x[2])
processed_data$label = sapply(strsplit(rownames(processed_data), "_"), function(x) x[1])
```

2- We will remove samples with NAs from all groups as we can't work on time points that have missing values at a specific time point for a patient.

```{r}
filtered_data = processed_data[!processed_data$sample_number %in% c(6,18,11,21,22,25),]
```


>> checking the NAs again 

```{r}
rowSums(is.na(filtered_data))
```

```{r}
head(filtered_data)
```


>> now let's work on checking the significance between the groups
    by extracting the metabolite values again and creating a vector with the group label

```{r}
metabolites <- filtered_data[, 10:(ncol(filtered_data)-2)] 

labels <- as.factor(filtered_data$label)   
```



>> we have 4 non-parametric groups so we will use kruskal-wallis test

```{r}
kruskal_results <- apply(metabolites, 2, function(metabolite) {
  kruskal_test <- kruskal.test(metabolite ~ labels) 
  return(c( "P_Value" = kruskal_test$p.value))
})
```



creating a df for the metabolites and their p.value

```{r}
kruskal_results_df <- as.data.frame(kruskal_results, row.names = names(kruskal_results))
colnames(kruskal_results_df) <- c( "P_Value")
```

>> subsetting the df with the significant proteins (p.value < 0.05)

```{r}
kruskal_significant_results <- subset(kruskal_results_df, P_Value < 0.05)
kruskal_significant_results
```

```{r}
dim(kruskal_significant_results)
```
we have 74 significant metabolite between the 4 groups 



>> now what about checking the signifcane between the time points itself
 
we need to create a new df for the 3 groups only  
 
```{r}
time_points_data <- filtered_data[filtered_data$label %in% c("D0", "D15", "D30"), ]

metabolites_time_points <- time_points_data[, 10:ncol(time_points_data)]

labels_time_points <- time_points_data$label
```


now we will melt the data with the sample number and the label columns we have created
>> each sample must has same number or name between the groups  

```{r}
long_data <- melt(metabolites_time_points)
head(long_data)
```


>> now let's test the significance between the 3 timepoints with Friedman test

```{r}
#Friedman test per metabolite
friedman_results <- by(long_data, ~ variable, function(data) {
  friedman_test <- friedman.test(value ~ label | sample_number, data = data)
  return(c("P_Value" = friedman_test$p.value))
})
```


```{r}
friedman_results_df <- as.data.frame(friedman_results, row.names = names(friedman_results))
colnames(friedman_results_df) <- c("P_Value")
```

>> as we did in kruskal, let's subset the significant proteins 

```{r}
friedman_significant_results <- subset(friedman_results_df, P_Value < 0.05)
```

```{r}
dim(friedman_significant_results)
```
there are 66 significant metabolite 




>> drawing a heatmap for friedman_significant_results using the raw values 
  1- extracting the significant metabolites from the data 
  2- extracting the used samples in the used groups (d0,d15,d30)
  3- creating a df for them 


```{r}
data = t(data)
significant_rows <- rownames(data) %in% rownames(friedman_significant_results)
significant_columns <- colnames(data) %in% colnames(t(metabolites_time_points))

significant_between_timepoints <- as.data.frame(data[significant_rows, significant_columns])
```

```{r}
dim(significant_between_timepoints)
```

```{r}
str(significant_between_timepoints)
```


```{r}
numeric_matrix <-  apply(significant_between_timepoints,2,as.numeric)
rownames(numeric_matrix) = rownames(significant_between_timepoints)
```

```{r}
heatmap.2(
  numeric_matrix,
  col = colorRampPalette(c("blue", "white", "red"))(200),  
  trace = "none",    
  scale = "row",
  dendrogram = "both",  
  main = "Significant metabolites Between Timepoints",
  density.info = "none", 
  labCol = colnames(numeric_matrix),  
  labRow = rownames(numeric_matrix)  
)
```






























